---
title: "Writing new MCMC samplers (or other nimbleFunctions)"
author: "Perry de Valpine"
date: "June 2020"
output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nimble)
```

Agenda
=====


Load Deer E. cervi example
=====
```{r load-DeerEcervi}
source(file.path("..", "examples", "DeerEcervi", "load_DeerEcervi.R"), 
       chdir = TRUE)
set.seed(123)
DEmodel1 <- nimbleModel(DEcode1,
                        constants = DEconstants,
                        data = list(Ecervi.01 = DeerEcervi$Ecervi.01),
                        inits = DEinits1())
```


Design of `nimble`'s MCMC system
=====

1. MCMC configuration object: Contains a list of sampler assignments, not actual samplers.
2. MCMC object: Contains a list of sampler objects.

Look at MCMC configuration object for just a few nodes
```{r}
mcmcConf <- configureMCMC(DEmodel1, nodes = 'farm.effect[1:3]')
class(mcmcConf)
ls(mcmcConf)
fe1_sampler <- mcmcConf$getSamplers("farm.effect[1]")[[1]]
class(fe1_sampler)
ls(fe1_sampler)
fe1_sampler$target
fe1_sampler$control
```


Two-stage evaluation in nimbleFunctions
=====
Say we want a nimbleFunction to calculate some nodes and their dependencies.

```{r}
calcDepsNF <- nimbleFunction(
  setup = function(model, nodes) {
    calcNodes <- model$getDependencies(nodes)
  },
  run = function() {
    ans <- model$calculate(calcNodes)
    return(ans)
    returnType(double())
  }
)
```

Two-stage evaluation in nimbleFunctions
=====
```{r}
calcDeps_farm1 <- calcDepsNF(DEmodel1, 'farm.effect[1]') ## first stage: setup code
calcDeps_farm1$run()
```

### Go to R to demonstrate uncompiled browsing/debugging.

More about nimbleFunctions
=====

- Without setup code, a `nimbleFunction` becomes an R function and a C++ function.
- With setup code, a `nimbleFunction` becomes an R reference class definition and a C++ class definition.

### nimbleFunction class definitions (i.e with setup code)

- `setup` is always run in R.
    - Typically one-time, high-level processing such as querying model structure.
- `run` and other methods can be run uncompiled (in R) or compiled (via C++).
    - Typically repeated "actual algorithm" calculations such as MCMC sampler updates.
- Any objects (e.g. `calcNodes` and `model`) in `setup` can be used in `run`.
    - Internally, these are automatically set up as class member data.

Step through setup and run code in uncompiled execution
=====
```{r eval}
calcDepsNF <- nimbleFunction(
  setup = function(model, nodes) {
    browser()
    calcNodes <- model$getDependencies(nodes)
  },
  run = function() {
    browser()
    ans <- model$calculate(calcNodes)
    return(ans)
    returnType(double())
  }
)
```

```{r eval=FALSE}
calcDeps_farm1 <- calcDepsNF(DEmodel1, 'farm.effect[1]') ## first stage: setup code
calcDeps_farm1$run()
```

A basic Random-Walk Metropolis-Hastings sampler
=====
```{r}
ourMH <- nimbleFunction(
  name = 'ourMH',
  contains = sampler_BASE,
  setup = function(model, mvSaved, target, control) {
    scale <- if(!is.null(control$scale)) control$scale else 1
    calcNodes <- model$getDependencies(target)
  },
  run = function() {
    currentValue <- model[[target]]
    currentLogProb <- model$getLogProb(calcNodes)
    proposalValue <- rnorm(1, mean = currentValue, sd = scale)
    model[[target]] <<- propValue
    proposalLogProb <- model$calculate(calcNodes)
    logAcceptanceRatio <- currentLogProb - proposalLogProb
    accept <- decide(logMHR)
    if(accept)
      copy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
    else     
      copy(from = mvSaved, to = model, row = 1, nodes = calcNodes, logProb = TRUE)
  },
  methods = list(
    reset = function() {}
  )
)
```

Rules for each sampler
=====

### setup function
  
- The four arguments, named exactly as shown, are required.  This allows `buildMCMC` to create any sampler correctly.

### run function

- The `mvSaved` ("modelValues saved") has a saved copy of all model variables and log probabilities
- Upon entry, `run()` can assume:
    - the model is fully calculated (so `getLogProb` and `calculateDiff` make sense).
    - `mvSaved` and the model are synchronized (have the same values).
- Upon exit, `run()` must ensure those conditions are met.
    - That way the next sampler can operator correctly.
- Between entry and exit, `run()` can manipulate the model in any way necessary.

### reset function

- To match the `sampler_BASE` definition, all samplers must have a `reset()` function.

Introduction to `modelValues`
=====

- 

Stepping through uncompiled execution
=====

