---
title: "Introduction to NIMBLE"
author: "Perry de Valpine"
date: "June 2020"
output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
has_ggplot2 <- require(ggplot2)
has_mcmcplots <- require(mcmcplots)
has_coda <- require(coda)
generate_original_results <- TRUE
```

# What Is NIMBLE?

- A framework for hierarchical statistical models and methods.
- A nearly drop-in alternative to WinBUGS, OpenBUGS and JAGS.
- An extension of the BUGS language for writing new functions and distributions.
- A configurable system for MCMC.
- A library of other methods.
- A model-generic programming system to write new analysis methods.
- Numerical Inference for statistical Models using Bayesian and Likelihood Estimation.

# First example: A generalized linear mixed model (GLMM)

* There are plenty of specialized GLMM packages.
* We start with a GLMM because it will be familiar to many of you.
* This example is from Zuur et al.
* E. cervi is a parasite of deer.  On each of 24 farms in Spain, multiple deer were sampled for E. cervi. 
* 826 total deer.
* Sex (M/F) and (centered) body Length are explanatory variables (fixed effects).
* Farm is a random effect.
* Response variable is presence (1) or absence (0) of the parasite E. cervi.

# GLMM example: Load the package
```{r}
library(nimble)
```

# GLMM example: Load data

```{r}
DeerEcervi <- read.table('DeerEcervi.txt', header = TRUE)
summary(DeerEcervi)

## Create presence/absence data from counts.
DeerEcervi$Ecervi.01 <- DeerEcervi$Ecervi
DeerEcervi$Ecervi.01[DeerEcervi$Ecervi>0] <- 1
## Center Length for better interpretation
DeerEcervi$cLength <- DeerEcervi$Length - mean(DeerEcervi$Length)
## Make a factor version of Sex for plotting
DeerEcervi$fSex <- factor(DeerEcervi$Sex)
## Make a factor and id version of Farm
DeerEcervi$fFarm <- factor(DeerEcervi$Farm)
DeerEcervi$farm.ids <- as.numeric(DeerEcervi$fFarm)
```

# GLMM example: Look at data

```{r eval=has_ggplot2}
ggplot(data = DeerEcervi, 
        mapping = aes(x = cLength, y = Ecervi.01, color = fSex)) + 
  geom_point() + 
  geom_jitter(width = 0, height = 0.1) + 
  facet_wrap(~Farm)
```

# GLMM example: Write the model code **in R**

```{r}
DEcode1 <- nimbleCode({
  for(i in 1:2) {
    # Priors for ntercepts and length coefficients for sex = 1,2
    sex.effect[i] ~ dnorm(0, sd = 1000)
    length.effect[i] ~ dnorm(0, sd = 1000)
  }
  
  # Priors for farm random effects and their standard deviation.
  farm.sd ~ dunif(0, 20)
  for(i in 1:num.farms) {
    farm.effect[i] ~ dnorm(0, sd = farm.sd)
  }
  
  # logit link and Bernoulli data probabilities
  for(i in 1:num.animals) {
    logit(disease.probability[i]) <- 
      sex.effect[ sex[i] ] +
      length.effect[ sex[i] ]*cLength[i] +
      farm.effect[ farm.ids[i] ]
    Ecervi.01[i] ~ dbern(disease.probability[i])
  }
})
```

GLMM Example: Steps to use NIMBLE:
===============

1. Build the model.  It is an R object.
2. Build the MCMC.

    - 2a. Configure the MCMC.
    - 2b. Customize the MCMC.
    - 2c. Build the MCMC.

3. Compile the model and MCMC.
4. Run the MCMC.
5. Extract the samples.

- `nimbleMCMC` does all of this at once.
- Steps 2a-2c can be combined if no customization is needed.
- `runMCMC` does steps 4-5

    - manages burn-in, multiple chains, and extracting samples.
    
GLMM Example: 1. Build the model
=====

```{r}
DEconstants <- list(num.farms = 24,
                    num.animals = 826,
                    cLength = DeerEcervi$cLength,
                    sex = DeerEcervi$Sex,
                    farm.ids = DeerEcervi$farm.ids)

DEmodel1 <- nimbleModel(DEcode1,
                        constants = DEconstants)
```

GLMM Example: 1b. Set data and inits
=====

These can be provided to `nimbleModel` or now:
```{r}
DEmodel1$setData(list(Ecervi.01 = DeerEcervi$Ecervi.01))
# This sets the values and *flags the nodes as data*.
DEinits1 <- function() {
  list(sex.effect = c(0, 0), 
       length.effect = c(0, 0),
       farm.sd = 1,
       farm.effect = rnorm(24, 0, 1) )       
}
set.seed(123)
DEmodel1$setInits(DEinits1())
```

GLMM Example: 2. Build the MCMC
=====
```{r}
DEmcmc1 <- buildMCMC(DEmodel1)
```

GLMM Example: 3. Compile the model and MCMC.
=====

This can be done in one step or two.  We'll use two.

```{r}
cDEmodel1 <- compileNimble(DEmodel1) 
# First call to compileNimble in a session is slower than later calls.
cDEmcmc1 <- compileNimble(DEmcmc1, project = DEmodel1)
```

GLMM Example: 4. Run the MCMC
=====
```{r}
cDEmcmc1$run(10000)
```

GLMM Example: 5. Extract the samples
=====
```{r}
samples <- as.matrix(cDEmcmc1$mvSamples)
```

GLMM Example: Look at results:
=====

There are many packages for summarizing and plotting MCMC samples.  NIMBLE does not try to re-invent these wheels.

1. `mcmcplots`

```{r eval=(has_mcmcplots & !generate_original_results)}
library(mcmcplots)
mcmcplot(samples, dir = ".", filename = "Ecervi_samples_mcmcplot")
```

```{r echo=FALSE, eval=generate_original_results}
library(mcmcplots)
mcmcplot(samples, dir = ".", filename = "orig_Ecervi_samples_mcmcplot")
```

Results that comes with these slides are [here](orig_Ecervi_samples_mcmcplot.html).

Results if you generated your own will be [here](Ecervi_samples_mcmcplot.html).

2. `coda`

```{r eval = (has_coda & !generate_original_results)}
library(coda)
pdf("Ecervi_samples_coda.pdf")
plot(as.mcmc(samples))
dev.off()
```

```{r echo = FALSE, eval = generate_original_results}
library(coda)
pdf("orig_Ecervi_samples_coda.pdf")
plot(as.mcmc(samples))
dev.off()
```

Results that comes with these slides are [here](orig_Ecervi_samples_coda.pdf).

Results if you generated your own will be [here](Ecervi_samples_coda.pdf).

MCMC workflow in NIMBLE
=====

Here is a summary of the MCMC workflow in NIMBLE.

[nimbleWorkflow.pdf](nimbleWorkflow.pdf)

GLMM Example: Doing it all at once with `nimbleMCMC`
=====

Start from:

- code
- constants + data
- inits

```{r}
set.seed(123)
DEdataAndConstants <- c(DEconstants, 
                        list(Ecervi.01 = DeerEcervi$Ecervi.01))
samples2 <- nimbleMCMC(DEcode1,
                       constants = DEdataAndConstants,
                       inits = DEinits1,
                       niter = 10000,
                       nburnin = 1000,
                       nchains = 2,
                       samplesAsCodaMCMC = TRUE)
summary(samples2) ## from coda
```

GLMM Example: Managing the run with `runMCMC`
=====

Start from:

- Compiled MCMC

```{r}
samples3 <- runMCMC(cDEmcmc1, 
                    niter = 10000,
                    nburnin = 1000,
                    nchains = 2,
                    samplesAsCodaMCMC = TRUE)
summary(samples3)
```

Other notes
=====

- `readBUGSmodel` will read BUGS/JAGS model code and variables from their standard file formats.

